### 2014-11-13 coordination.tex
Add locking section.
Fix more typos.

The SSSP program present in Fig.~\ref{code:shortest_path_program} is concise and
declarative but its performance depends on the order in which nodes are
executed. If nodes with greater distances are prioritized over other nodes, the
program will generate more \texttt{relax} facts since it will take longer to
reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
clear that the best scheduling is the following: \texttt{@1},
\texttt{@3}, \texttt{@2} and then \texttt{@4}, where only 4 \texttt{relax}
facts are generated. If we had decided to process nodes in order
\texttt{@1}, \texttt{@2}, \texttt{@4}, \texttt{@3}, \texttt{@4},
\texttt{@2}, then 6 \texttt{relax} facts would have been generated.
The optimal solution for SSSP is to schedule the node with the
shortest distance, which is essentially the Dijkstra shortest path
algorithm~\cite{Dijkstra}. Note how it is possible to change the nature of
the algorithm by simply changing the order of node computation, but still
retain the declarative nature of the program.

@@

\begin{itemize}
   \item \code{set-priority(node A, float F)}: This sets the
   temporary priority of \texttt{A} to \texttt{F}. If \texttt{A} has a priority
   \texttt{F'}, we only change the priority if \texttt{F} is higher. The programmer
   can decide if priorities are to be ordered in ascending or descending order.
   \item \code{add-priority(node A, float F)}: Increases,
   temporarily, the priority of node \texttt{A} by \texttt{F}.
   \item \code{schedule-next(node A)}: Changes the temporary priority of node
   \texttt{A} to be 1 unit higher than the current highest priority.
   \item \code{set-default-priority(node, float)}: Sets the default
   priority of the node.
   \item \code{stop-program(node A)}: Immediately stops the
   execution of the whole program.
\end{itemize}

@@

\subsubsection{Partitioning facts}

We provide several coordination facts for dealing with node partitioning among
the available threads executing the program. In terms of action
facts, we have the following:

\begin{itemize}
   \item \code{set-cpu(node A, int C)}: Moves node \texttt{A} to thread
   \texttt{C}.
   \item \code{set-affinity(node A, node B)}: Places node \texttt{B} in
   the thread of node \texttt{B}.
   \item \code{set-moving(node A)}: Allows node \texttt{A} to move freely
   between threads.
   \item \code{set-static(node A)}: Forces node \texttt{A} to stay in the
   same thread indefinitely.
\end{itemize}

For sensing facts, we have the following set of coordination facts:

\begin{itemize}
   \item \code{cpu-id(node A, node B, int C)}: Store at node \texttt{A} which
   thread, \texttt{C}, \texttt{B} is actually running on.
   \item \code{moving(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is allowed
   to move between threads.
   \item \code{static(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is not
   allowed to move between threads.
\end{itemize}

@@

\begin{description}
   \item[\texttt{priority @order ORDER.}] \texttt{ORDER} can be either \texttt{asc} or \texttt{desc}. This defines if node's are to be selected by the smallest or the greatest priority, respectively.
   \item[\texttt{priority @initial P.}] The \texttt{initial} statement informs the runtime system that all nodes must start with priority $P$. Alternatively, the programmer can define an \texttt{set-priority(A, P)} axiom.
   \item[\texttt{priority @static.}] The \texttt{static} priority tells the runtime system that the partition of nodes among workers is to be used until the end of program. 
\end{description}

>>>>

The SSSP program is concise and declarative but its performance depends on the order in which nodes are
executed. If nodes with greater distances are prioritized over other nodes, the
program will generate more \texttt{relax} facts since it will take longer to
reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
clear that the best scheduling is the following: \texttt{@1},
\texttt{@3}, \texttt{@2} and then \texttt{@4}, where only 4 \texttt{relax}
facts are generated. If we had decided to process nodes in order
\texttt{@1}, \texttt{@2}, \texttt{@4}, \texttt{@3}, \texttt{@4},
\texttt{@2}, then 6 \texttt{relax} facts would have been generated.
The optimal solution for SSSP is to schedule the node with the
shortest distance, which is essentially the Dijkstra shortest path
algorithm~\cite{Dijkstra}. Note how it is possible to change the nature of
the algorithm by simply changing the order of node computation, but still
retain the declarative nature of the program.

@@

\begin{tightitemize}
   \item \code{set-priority(node A, float F)}: This sets the
   temporary priority of \texttt{A} to \texttt{F}. If \texttt{A} has a priority
   \texttt{F'}, we only change the priority if \texttt{F} is higher. The programmer
   can decide if priorities are to be ordered in ascending or descending order.
   \item \code{add-priority(node A, float F)}: Increases,
   temporarily, the priority of node \texttt{A} by \texttt{F}.
   \item \code{remove-priority(node A)}: Removes the temporary priority from node
   \texttt{A}.
   \item \code{schedule-next(node A)}: Changes the temporary priority of node
   \texttt{A} to be 1 unit higher than the current highest priority.
   \item \code{set-default-priority(node, float)}: Sets the default
   priority of the node.
   \item \code{stop-program(node A)}: Immediately stops the
   execution of the whole program.
\end{tightitemize}

@@

\subsection{Partitioning facts}

We provide several coordination facts for dealing with node partitioning among
the running threads. In terms of action facts, we have the following:

\begin{tightitemize}
   \item \code{set-cpu(node A, int T)}: Moves node \texttt{A} to thread
   \texttt{T}.
   \item \code{set-affinity(node A, node B)}: Places node \texttt{B} in
   the thread of node \texttt{A}.
   \item \code{set-moving(node A)}: Allows node \texttt{A} to move freely
   between threads.
   \item \code{set-static(node A)}: Forces node \texttt{A} to stay in the
   same thread indefinitely.
\end{tightitemize}

For partitioning facts, we have the following set of sensing facts:

\begin{tightitemize}
   \item \code{cpu-id(node A, node B, int T)}: Store at node \texttt{A} which
   thread \texttt{T}, \texttt{B} is actually running on.
   \item \code{moving(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is allowed
   to move between threads.
   \item \code{static(node A, node B)}: Fact available at node \texttt{A} if \texttt{B} is not
   allowed to move between threads.
\end{tightitemize}

@@

\begin{tightdescription}
   \item[\texttt{priority @order ORDER.}] \texttt{ORDER} can be either \texttt{asc} or \texttt{desc}. This defines if node's are to be selected by the smallest or the greatest priority, respectively.
   \item[\texttt{priority @initial P.}] The \texttt{initial} statement informs the runtime system that all nodes must start with priority $P$. Alternatively, the programmer can define an \texttt{set-priority(A, P)} axiom.
   \item[\texttt{priority @static.}] The \texttt{static} priority tells the runtime system that the partition of nodes among workers is to be used until the end of program. 
\end{tightdescription}

### 2014-11-13 heat-transfer.tex
small space savings
Fix more typos.

Fig.~\ref{results:ht} presents the scalability results for the uncoordinated
and coordinated version. The dataset used is a square grid with an inner square
with high heat nodes. With 1 thread, there's a 33\% reduction in run time, while
for 16 threads, there's a 18\% reduction.

To further improve locality, we split the second rule to avoid sending
small $\delta$ values if the target node is in another thread
(Fig.~\ref{code:ht_better}). The \textbf{Local-Only/Coordinated} line
in Fig.~\ref{results:ht} shows that this performs better due to
improved locality, exhibiting linear scalability over the regular
version.  However, this come at the price of increased errors in the
computed heat values.

>>>>

Fig.~\ref{results:ht} presents the scalability results for the regular
and coordinated version. The dataset used is a square grid with an inner square
with high heat nodes. Comparing the coordinated version with the regular
version, with 1 thread there is a 33\% reduction in run time, while
for 16 threads there is a 18\% reduction.

To further improve locality, we have split the second rule to avoid sending
small $\delta$ values if the target node is in another thread
(Fig.~\ref{code:ht_better}). The \textbf{Local-Only/Regular} line
in Fig.~\ref{results:ht} shows that this has the best performance.  However, this
come at the price of increased errors in the computed heat values.

### 2014-11-13 implementation.tex
Fix more typos.

In the paper~\cite{cruz-ppdp14}, the LM virtual machine was measured and
compared against implementations of similar programs but written in other
languages. For instance, the LBP program was found to be around 2 times slower
than GraphLab, while N Queens program was found to be around 10 to 15 slower
than a sequential C implementation. Some of those programs were compared against
a Python implementations and LM faired fairly well. We measured
the impact of the coordination mechanisms implemented for the CLM
virtual machine on regular LM programs and saw very little to no degration.
Since we also performed general performance improvements, programs such as
PageRank and Greedy Graph Coloring programs now perform better than before.

>>>>

In the paper~\cite{cruz-ppdp14}, the LM virtual machine was measured and
compared against implementations of similar programs but written in other
languages. For instance, the Loopy Belief Propgation program was found to be around 2 times slower
than GraphLab, while the N Queens program was found to be around 10 to 15 slower
than a sequential C implementation. Some of those programs were also compared against
a Python implementations and LM faired fairly well. We measured
the impact of the coordination mechanisms implemented for the CLM
virtual machine on regular LM programs and saw very little to no degration.
Since we also performed general performance improvements, programs such as
PageRank and Greedy Graph Coloring now perform slightly better than before.

### 2014-11-13 locks.tex
Fill percent fail.
Fix more typos.

\begin{toptab}
\footnotesize\begin{tabular}{ | l || c | c | c | c | c |}
\hline
\multirow{2}{*}{\textbf{Program}} & \multicolumn{5}{c |}{\textbf{Queue
   Instructions}} \\ \cline{2-6}
& 1 & 2 & 4 & 8 & 16 \\ \hline
\hline
SSSP - Regular & 199K & 201K & 205K & 196K & 207K \\ \hline
SSSP - Coordinated & 964K & 938K & 870K & 886K & 963K \\ \hline
SSSP - Buffered & 1M & 1M & 1M & 1M & 1M \\ \hline
\hline
HT - Regular & 25M & 32M & 34M & 34M & 36M \\ \hline
HT - Coordinated & 51M & 55M & 85M & 80M & 72M \\ \hline
HT - Local-Only & 51M & 55M & 79M & 74M & 68M \\ \hline
\hline
LBP - Regular & 15M & 17M & 16M & 11M & 8M \\ \hline
SBP - Coordinated & 96M & 97M & 87M & 83M & 77M \\ \hline
\hline
MiniMax - Regular & 16M & 19M & 19M & 19M & 20M \\ \hline
MiniMax - Coordinated & 51M & 51M & 51M & 51M & 52M \\ \hline
\hline
N Queens - Regular & 8K & 15K & 16K & 17K & 20K \\ \hline
N Queens - Coordinated & 57M & 24M & 675K & 145K & 46K \\ \hline
N Queens - Static & 57M & 44M & 1M & 1M & 151K \\ \hline
\hline
\end{tabular}
\vspace*{.5ex}
\scap{queue_instructions}{Number of queue \scare{instructions} per program. An
   operation to a regular queue counts as 1 instruction, while an operation to
   a priority queue costs the number of
   \emph{percolate-up}/\emph{percolate-down} steps performed.}
\end{toptab}

Coordination introduces overhead in two major ways.  First, in
manipulating the priority queues which require operations on a min
heap.  Second, in potentially requiring more lock operations as we
move nodes between queues and within the priority queue.  As we show
here, this overhead is minimal and the resulting improved schedules
and improved locality more than make up for the overhead.

As can be seen in Table~\ref{queue_instructions} all the coordinated
versions of the programs require significantly more queue operations.
The data for this table comes from recording, during execution, the
number of normal queue operations executed (each costs 1), plus the
number of \emph{percolate-up}/\emph{percolate-down} operations
executed for each manipuilation of the priority queue.  As expected,
programs which rely on prioritizing the order of execution to obtain
better asymptotic complexity (e.g., SSSP) improve in overall running
time at the cost of more queue operations.  In all cases, coordination
adds a significant number of total queue operations, but the resulting
overhead is more than compensated for by an improved schedule.

@@

We next investigate the effect of coordination on the syncronization
overhead that comes froms from the use of more locks.  We measured the
\emph{basic locks}, the number of locks acquired
in the original LM virtual machine and \emph{coordination locks}, the
number of locks acquired to perform coordination operations.  We also
measured how often \texttt{lock()} operations that fail to acquired
the lock on the first attempt.  Lock failures happen so infrequently
(less than X\% \snote{please fill this in} of the time) that they
cannot be seen on the figures.

Fig.~\ref{results:sssp_uspowergrid} shows the number of locks acquired
for the SSSP program.  The first bar in each group shows the number of
locks acquired for the uncoordinated version of the program.  The
second bar shows the regular and coordination locks acquired for SSSP
when every newly derived fact and priority operation is carried out
immediately.  This leads to a significant increase in total locks
acquired often because the priority of a node is change many times
before it is operated on.  The third bar shows the effect of buffering
priority operations before applying them.  This significantly reduces
the lock overhead, but as can be seen in
Table~\ref{queue_instructions} increase the number of queue operations
slightly.  This slight increase comes from the fact the buffering
delays the change in priorities and some longer paths can sometimes be
investigated unnecessarily.

Fig.~\ref{locks:HT} shows the locking behavior of the heat transfer
program.  The behavior is representative of the other programs.  There
is drop in the total number of locks acquired and coordination locks
represent a small fraction of the total locks acquired.  The locality
coordination further reduces locking by keeping more derivation local
to a thread.

Overall, the additional overhead introduced by supporting coordination
is more than offset by improved schedules and increased locality.
While there are more locks to manage the total cost of lock operations
does not change significantly (and is often reduced) when coordination
is added.  And, although the priority coordination significantly
increases the cost of queueing it is offset by reducing the number of
total facts derived.

>>>>

\begin{toptab}
\footnotesize\begin{tabular}{ | l || c | c | c | c | c |}
\hline
\multirow{2}{*}{\textbf{Program}} & \multicolumn{5}{c |}{\textbf{Queue
   Instructions}} \\ \cline{2-6}
& 1 & 2 & 4 & 8 & 16 \\ \hline
\hline
SSSP - Regular & 199K & 201K & 205K & 196K & 207K \\
SSSP - Coordinated & 964K & 938K & 870K & 886K & 963K \\
SSSP - Buffered & 1M & 1M & 1M & 1M & 1M \\
\hline
HT - Regular & 25M & 32M & 34M & 34M & 36M \\
HT - Coordinated & 51M & 55M & 85M & 80M & 72M \\
HT - Local-Only & 51M & 55M & 79M & 74M & 68M \\
\hline
LBP - Regular & 15M & 17M & 16M & 11M & 8M \\
SBP - Coordinated & 96M & 97M & 87M & 83M & 77M \\
\hline
MiniMax - Regular & 16M & 19M & 19M & 19M & 20M \\
MiniMax - Coordinated & 51M & 51M & 51M & 51M & 52M \\
\hline
N Queens - Regular & 8K & 15K & 16K & 17K & 20K \\
N Queens - Coordinated & 57M & 24M & 675K & 145K & 46K \\
N Queens - Static & 57M & 44M & 1M & 1M & 151K \\
\hline
\end{tabular}
\vspace*{.5ex}
\scap{queue_instructions}{Number of queue \scare{instructions} per program. An
   operation to a regular queue counts as 1 instruction, while an operation to
   a priority queue costs the number of
   \emph{percolate-up}/\emph{percolate-down} steps performed.}
\end{toptab}

Coordination introduces overhead in two major ways.  First, in
manipulating the priority queues which require operations on a min
heap.  Second, in potentially requiring more lock operations as we
move nodes between queues and within the priority queue.  As we show
next, this overhead is minimal and is compensated by the resulting improved schedules
and memory locality that more than make up for the overhead.

As can be seen in Table~\ref{queue_instructions} all the coordinated
versions of the programs require significantly more queue operations.
The data for this table comes from recording, during execution, the
number of normal queue operations executed (each costs 1), plus the
number of \emph{percolate-up}/\emph{percolate-down} operations
executed for each manipulation of the priority queue.  As expected,
programs which rely on prioritizing the order of execution to obtain
better asymptotic complexity (e.g., SSSP) improve in overall running
time at the cost of more queue operations.  In all cases, coordination
adds a significant number of total queue operations, but the resulting
overhead is more than compensated for by an improved schedule.

@@

We next investigate the effect of coordination on the syncronization
overhead that comes from the use of more locks.  We measured the
\emph{basic locks}, the number of locks acquired
in the original LM virtual machine and \emph{coordination locks}, the
number of locks acquired to perform coordination operations.  We also
measured how often \texttt{lock()} operations that fail to acquired
the lock on the first attempt.  Lock failures happen so infrequently
(less than 1\% of the time) that they cannot be seen on the figures.

Fig.~\ref{results:sssp_uspowergrid} shows the number of locks acquired
for the SSSP program.  The first bar in each group shows the number of
locks acquired for the regular version of the program.  The
second bar shows the basic and coordination locks acquired for the coordinated
version of SSSP when every newly derived fact and priority operation is carried out
immediately.  This leads to a significant increase in total locks
acquired often because the priority of a node is change many times
before it is operated on.  The third bar shows the effect of buffering
priority operations before applying them.  This significantly reduces
the lock overhead, but as can be seen in
Table~\ref{queue_instructions}, increases the number of queue operations
slightly.  This slight increase comes from the fact that buffering
delays the change in priorities and some longer paths can sometimes be
investigated unnecessarily.

Fig.~\ref{locks:HT} shows the locking behavior of the heat transfer
program.  The behavior is representative of the other programs. In general, there
is a drop in the total number of locks acquired and coordination locks
represent a small fraction of the total number of locks acquired.  The \textbf{Local-Only}
version further reduces locking by keeping more derivations local
to a thread.

Overall, the additional overhead introduced by supporting coordination
is more than offset by improved schedules and increased locality.
While there are more locks to manage the total cost of lock operations,
the numbers do not change significantly (and is often reduced) when coordination
is added.  And, although the priority coordination significantly
increases the cost of queueing it is offset by reducing the number of
total facts derived.

### 2014-11-13 minimax.tex
Fix more typos.

The code in Fig.~\ref{minimax:check-end} deals with expanding the tree.
The first three rules (lines 1-10) deal
with the case where no children nodes are created and the last three rules
(12-29) deal with the cases for creating new nodes.
The rule in lines 12-25 generates new nodes using the
\texttt{exists} language construct, which creates a child node 
\texttt{B}. We link \texttt{B} with its parent (\texttt{parent(B, A)})
and kick start the expansion of that node \texttt{B} by adding a \texttt{play}
fact.

@@

With coordination, we set the priority of a node to be its depth
so the tree is expanded in a depth-first fashion (lines 15 and 22),
leading to a memory complexity of $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads.
Since threads prioritize deeper nodes, the scores of the first leaves are immediately
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

>>>>

The code in Fig.~\ref{minimax:check-end} deals with the tree expansion process.
The first three rules (lines 1-10) deal
with the case where no children nodes are created and the last three rules
(12-29) deal with the cases for that create new nodes. In particular, the two
rules in lines 12-25 generate new nodes using the
\texttt{exists} language construct, which creates a child node 
\texttt{B}. We link \texttt{B} with its parent (\texttt{parent(B, A)})
and kick start the expansion of that node \texttt{B} by adding a \texttt{play}
fact.

@@

With coordination, we set the priority of a node to be its depth (lines 15 and
22) so the tree is expanded in a depth-first fashion,
leading to a memory complexity of $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads.
Since threads prioritize deeper nodes, the scores of the first leaves are immediately
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

### 2014-11-13 nqueens.tex
Fix more typos.

The N-Queens puzzle is the problem of placing N chess queens on an NxN
chessboard so that no pair of two queens attack each
other~\cite{8queens}. The challenge of finding all the
distinct solutions to this problem is a good benchmark in designing
parallel algorithms.

@@

Since computation goes from the top row to the bottom row, not all
placements of nodes to threads performane equally. This is especially
true because the bottom rows tend to perform the most work. The best
placement is then to split the board vertically so each thread gets
the same number of columns with axiom \texttt{set-cpu(A, vertical(X, Y)}, where
\texttt{X} and \texttt{Y} are the coordinates of the board. Our experiments show that, on a shared
memory system, it does not matter much if we start with a bad
placement since node stealing overcomes those issues by load balancing
dynamically.  However, the N Queens program incrementally builds and shares lists representing
valid board states that are transmitted from top to bottom.
In order to improve memory locality we should then manipulate board
states that share a significant number of elements because each board state
needs to be iterated before extended with a new position.
To accomplish this, we used coordination axioms \texttt{set-default-priority(A,
X)}, where \texttt{X} is the row number (0 is the top row) in order to
prioritize the evaluation of squares closer to the bottom of the board.

Experimental results are presented in Fig.~\ref{results:nqueens}.  We
first ran the coordinated program using dynamic partitioning (with
work stealing) and saw some performance improvements
(\textbf{Coordinated}). However, using dynamic partitioning is bad for
memory locality since threads will often steal nodes.  We then
added \texttt{set-static(A)} as a coordination axiom and saw
performance improvement (\textbf{Static}), especially when the board
columns are perfectly partitioned by the number of threads (in our
case 13 threads). We also analyzed N Queens execution with Valgrind's
CacheGrind~\ref{Nethercote:2007:VFH:1250734.1250746} and found that the
\texttt{Static} version has half the number of data caches misses (last level)
of the \texttt{Regular} version and a slight more than half of the
\texttt{Coordinated} version. This validates our hypothesis about the
improvement seen in the \texttt{Static} version.

>>>>

The N-Queens puzzle is the problem of placing N chess queens on an NxN
chessboard so that no pair of two queens attack each
other~\cite{8queens}. The challenge of finding all the
distinct solutions to this problem is a well-known benchmark in designing
parallel algorithms.

@@

Since computation goes from the top row to the bottom row, not all
placements of nodes to threads perform equally. This is especially
true because the bottom rows tend to perform the most work. The best
placement is then to split the board vertically so each thread gets
the same number of columns with axiom \texttt{set-cpu(A, vertical(X, Y)}, where
\texttt{X} and \texttt{Y} are the coordinates of the board. Our experiments show that, on a shared
memory system, it does not matter much if we start with a bad
placement since node stealing overcomes those issues by load balancing
dynamically.  However, the N Queens program incrementally builds and shares lists representing
valid board states that are transmitted from top to bottom.
In order to improve memory locality we should then manipulate board
states that share a significant number of elements because each board state
needs to be iterated before extended with a new position.
To accomplish this, we used coordination axioms \texttt{set-default-priority(A,
X)}, where \texttt{X} is the row number (0 is the top row) in order to
prioritize the evaluation of squares closer to the bottom of the board.

Experimental results are presented in Fig.~\ref{results:nqueens}.  We
first ran the coordinated program using dynamic partitioning (with
work stealing) and saw some performance improvements
(\textbf{Coordinated}). However, using dynamic partitioning is bad for
memory locality since threads will often steal nodes.  We then
added \texttt{set-static(A)} as a coordination axiom and saw
performance improvement (\textbf{Static}), especially when the board
columns are perfectly partitioned by the number of threads (in our
case 13 threads). We also analyzed N Queens execution with Valgrind's
CacheGrind~\cite{Nethercote:2007:VFH:1250734.1250746} and found that the
\texttt{Static} version has half the number of data caches misses (last level)
of the \texttt{Regular} version and a slight more than half of the
\texttt{Coordinated} version. This validates our hypothesis about the
improvement seen in the \texttt{Static} version.

### 2014-11-13 splash-bp.tex
removed redundant info about graphlab code
Fix more typos.

Randomized and approximation algorithms can obtain significant benefits from
coordination directives because their inherent non-determinism can be harnessed
to evaluate rules in different orders.
An example of such program is Loopy Belief Propagation.
Loopy Belief Propagation~\cite{Murphy99loopybelief} (LBP) is an approximate inference algorithm
used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
where nodes exchange messages with their immediate neighbors and apply some computations to the messages
received.

LBP is an algorithm that maps very well to the graph based model of CLM. In its
original form, the belief values of nodes are computed by synchronous iterations.
LBP offers more concurrency when belief values are computed asynchronously
leading to faster convergence. For this, every node keeps track of all messages
sent/received and recomputes the belief using partial information from neighbor
nodes. It is then possible to prioritize the computation of beliefs when a
neighbor's belief changes significantly.

@@

\begin{tightdescription}
   \item[Tree building]: Each node has a \texttt{inactive} fact that is used to
   start the tree building process. When the highest priority node is picked, a
   \texttt{tree} is created that will navigate through the tree. In lines 15-21,
   we use an \emph{aggregate}~\cite{cruz-iclp14} to gather all the neighbor
   nodes that have a positive priority (due to a new belief update) and are in the
   same thread. Nodes are collected into list \texttt{L} (line 21) and
   appended to list \texttt{Next}.

   \item[First phase]: In the
   third rule (lines 11-12), when the number of nodes of the tree reaches a
   certain limit, a \texttt{first-phase} is generated to update the beliefs of
   all nodes in the tree, starting from the leaves and ending at the root As the
   nodes are updated, an \texttt{update} fact is derived to update the belief
   values (line 35).

@@

// SECOND PHASE
second-phase(A, [], _)
   -o *underline{set-priority(A, 0.0)}, inactive(A), update(A).
second-phase(A, [A], Back)
   -o update(A), inactive(Back),
      inactive(A), *underline{set-priority(A, 0.0)}.
second-phase(A, [A, B | Next], Back)
   -o update(A), inactive(Back), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
\end{Verbatim}
  \scap{code:sbp}{Coordination code for the Splash Belief Propagation program. CLM needs
     50 lines of rules to implement splash trees, while GraphLab needs a
     total of 350 lines of C++ to implement the same functionality.  Note
     that when a linear fact is prefixed by \texttt{@}, it indicates that the
     fact is going to be re-derived.}
\end{topfig}
\normalsize

@@

In this program, coordination assumes a far more important role than
we have seen before. Coordination rules fully drive the behavior of
the algorithm and although the final result of the algorithm is
statistically identical to the original algorithm, SBP works very
differently than LBP.  SBP is also implemented in
GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine
algorithms.  GraphLab provides the splash scheduler as part its the
framework. It is 350 lines of C++ code.  With our coordination facts,
it is possible to create the necessary scheduling with only 12 rules.

>>>>

Randomized and approximation algorithms can obtain significant benefits from
coordination directives because their inherent non-determinism can be harnessed
to evaluate rules in different orders.
An example of such program is Loopy Belief Propagation~(LBP).
LBP~\cite{Murphy99loopybelief} is an approximate inference algorithm
used in graphical models with cycles. In its essence, LBP is a sum-product message passing algorithm
where nodes exchange messages with their immediate neighbors and apply some computations to the messages
received.

LBP maps very well to the graph based model of CLM. In its
original form, the belief values of nodes are computed by synchronous iterations.
LBP offers more concurrency when belief values are computed asynchronously
leading to faster convergence. For this, every node keeps track of all messages
sent/received and recomputes the belief using partial information from neighbor
nodes. It is then possible to prioritize the computation of beliefs when a
neighbor's belief changes significantly.

@@

\begin{tightdescription}
   \item[Tree building]: Each node has a \texttt{inactive} fact that is used to
   start the tree building process. When the highest priority node is picked, a
   \texttt{tree} is created that will navigate through the tree. In lines 15-21,
   we use an \emph{aggregate}~\cite{cruz-iclp14} to gather all the neighbor
   nodes that have a positive priority (due to a new belief update) and are in the
   same thread. Nodes are collected into list \texttt{L} and
   appended to list \texttt{Next} (line 21).

   \item[First phase]: When the number of nodes in the tree reaches a certain
   limit, a \texttt{first-phase} (lines 11-12) is generated to update the beliefs of
   all nodes in the tree. As the nodes are updated, starting from the leaves and
   ending at the root, an \texttt{update} fact is derived to update the belief
   values (line 35).

@@

// SECOND PHASE
second-phase(A, [], _)
   -o *underline{remove-priority(A)}, inactive(A), update(A).
second-phase(A, [A], Back)
   -o update(A), inactive(Back),
      inactive(A), *underline{remove-priority(A)}.
second-phase(A, [A, B | Next], Back)
   -o update(A), inactive(Back), *underline{schedule-next(B)},
      second-phase(B, [B | Next], A).
\end{Verbatim}
  \scap{code:sbp}{Coordination code for the Splash Belief Propagation program. 
%CLM needs 50 lines of rules to implement splash trees, while GraphLab needs a total of 350 lines of C++ to implement the same functionality.
  Note
     that when a linear fact is prefixed by \texttt{@}, it indicates that the
     fact is going to be re-derived.}
\end{topfig}
\normalsize

@@

In this program, coordination assumes a far more important role than
we have seen before. Coordination rules fully drive the behavior of
the algorithm and although the final result of the algorithm is
statistically identical to the original algorithm, SBP works very
differently than LBP.  SBP is also implemented in
GraphLab~\cite{GraphLab2010}, a C++ framework for writing machine
algorithms.  GraphLab provides the splash scheduler as part of its the
framework. It includes 350 lines of C++ code.  With our coordination facts,
it is possible to create the necessary scheduling with only 12 rules.

### 2014-11-13 sssp.tex
Fix more typos.

Here we add coordination to the SSSP program described in \sectref{sect:ssspex}.
The coordinated version of the
SSSP~(Fig.~\ref{code:shortest_path_program_coord}) uses the coordination fact
\texttt{set-priority} (line 14). We also use a global program directive to order
priorities in ascending order (line 5).

@@

The most interesting property of the SSSP program presented in
Fig.~\ref{code:shortest_path_program_coord} is that it remains provably correct,
although it applies rules using smarter ordering. The derivation of
\texttt{set-priority} does not change the behavior of the logical rules and the
code remains declarative.

Fig.~\ref{results:sssp_uspowergrid} shows experimental results when the
program computes the SSSP for 20\% of the nodes of the graph.
There are some situations where unnecessary facts are propagated
because although the shortest distance is selected locally, sub-optimal distances may be
propagated because many SSSP distances are computed at the same time.
However, we see a reduction of over 50\% in the number of
derived facts when using the coordinated version \textbf{C} over the regular
version \textbf{R}. The results also show that the coordinated
version using 16 threads is 1.3 times faster than the regular version using 16
threads while still showing good scaling.

>>>>

We start by adding coordination to the SSSP program described before.
The coordinated version of the
SSSP~(Fig.~\ref{code:shortest_path_program_coord}) uses the coordination fact
\texttt{set-priority} (line 14) and a global program directive to order
priorities in ascending order (line 5).

@@

The most interesting property of the SSSP program presented in
Fig.~\ref{code:shortest_path_program_coord} is that it remains provably correct,
although it applies rules using smarter ordering and the code remains
declarative. The derivation of \texttt{set-priority} does not change the
behavior of the logical rules.

Fig.~\ref{results:sssp_uspowergrid} shows experimental results when the
program computes the SSSP for 20\% of the nodes of the graph.
There are some situations where unnecessary facts are propagated
because although the shortest distance is selected locally, sub-optimal distances may be
propagated because many SSSP distances are computed at the same time.
However, we see a reduction of over 40\% in the number of
derived facts when using the coordinated version \textbf{C} over the regular
version \textbf{R}. The results also show that the coordinated
version using 16 threads is 1.3 times faster than the regular version using 16
threads while still showing good scaling.
